Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-26 icfp.tex
reverse binder swap muito bonito; sec 4 pass

When the scrutinee is not in WHNF, it is not statically known which resources
will have been consumed after evaluating said scrutinee, or which
are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
a \rightarrow e\}}$ where $f$ is defined as $f = K$:
\begin{enumerate}

@@

    \item On the other hand, since evaluation to WHNF does not guarantee all
    resources are consumed (unless the result matches an
    unrestricted pattern), we must ensure that any resources that transfer from the
    scrutinee to the alternative are ultimately consumed (in the example,
    $x$ is not consumed by evaluating the scrutinee).

@@

We develop the metatheory of Linear Core by first 
presenting the operational semantics of $\lambda^\pi_\D$,
consisting of a lazy natural semantics in the
style of Launchbury~\cite{10.1145/158511.158618}.
This semantics conveys the expected behavior of an implementation of
lazy evaluation and is agnostic to any linearity information.
Such a semantics is ill-suited for reasoning about
linearity and so we develop an \emph{instrumented} linearity-aware semantics that is
enriched with sufficient information to allow us to establish type
safety, showing that linearity is preserved. We show that the natural
and instrumented semantics are bisimilar for well-typed terms and so
derive type safety in the natural semantics
(Section~\ref{sec:typesafe}). Finally, we show that multiple optimising
transformations such as those performed by GHC on (Linear) Core are
type preserving
(Section~\ref{sec:optimisations-preserve-types-meta}), and thus
linearity preserving.

@@

The Launchbury-style natural semantics, presented in Figure~\ref{fig:opsem}, captures
standard lazy evaluation while ignoring linearity information.
% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide. 
%
%Both semantics are
The natural semantics is equipped with an evaluation environment that
maps variables to expressions and is
mutated in order to express evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders (see Appendix~\ref{app:opsem}).
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.

@@

  The natural semantics is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, $\Theta$ and $\Theta'$ are evaluation environments with bindings of the
form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
variable $x$ of the given type. The rules are standard, augmenting the
runtime environment with let-bound expressions which are evaluated
(only once) if the let-bound variable is forced.

@@

While the natural semantics is a mostly direct model of an implementation
of lazy evaluation, it is inconvenient for formal reasoning about
well-typed linear terms.
%
We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a
an \emph{instrumented} operational semantics in which linear variables are
erased from 
the runtime environment once they are forced, ensuring that any term violating linearity
will result in a stuck state. The semantics is also type-aware and
carries with it sufficient data to reconstruct typing derivations for
the purposes of showing type safety.

The semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.

@@

\todo{UPDATE}
Finally, we discuss the special case of scrutinizing a
single variable $x$, e.g.:
$
\lambda x.~\ccase{x}{\_ \to x}
$. 
It might seem as though this function is linear:
%begin{itemize}
% \item
if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact;
% \item
if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
$x$ returns the result of evaluating $x$ that still had to be consumed exactly
once.
%\end{itemize}
However, in practice, this program's linearity can only be determined by
further analysis on the non-strict evaluation strategy. If linear
function applications are $\beta$-reduced \emph{call-by-name}
(often desirable, as linear functions are known to use their argument exactly
once)
% and thus call-by-name reduction will definitely not duplicate computation
while the example function is considered linear, then such an application might
duplicate linear resources during evaluation. For instance:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
\Longrightarrow_{CBN}
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
On the other hand, if linear function applications are reduced
\emph{call-by-need}, the above example would instead be semantically linear
because the result of freeing $x$ would be bound to some $y$ first.
%
The system we present in the next section does accept this function
scrutinizing a variable as linear, as our operational semantics are
call-by-need.
This interaction between evaluation and linearity is considered specifically in the
reverse binder swap transformation discussed in
Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}

Reverse binder swap is dual to the binder swap transformation, substituting
occurrences of the case binder $z$ by the scrutinee in the special case when
the scrutinee is a variable $x$.
%
% his is dual to what the binder swap transformation
% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
By using the scrutinee $x$ instead of the case binder, we might be able to
float out expressions from the alternative using the case binder. For example,
we might float an expensive computation involving $z$ out of the case
alternative, where $z$ is now out of scope:

{\small\[
\begin{array}{l}
(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
\Longrightarrow_\textrm{Reverse binder swap}\\
(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
\Longrightarrow_\textrm{Commuting case-let}\\
\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
\end{array}
\]}

\noindent If $go$ is compiled into a loop, we only compute
$\mathit{expensive}~x$ once instead of in every loop iteration.

Even though GHC applies the reverse binder swap transformation in the
Core-to-Core passes, \emph{this optimisation violates linearity} when considered
as a transformation \emph{on Linear Core programs} because the result (second program
above) is ill-typed -- the resource $x$ of the scrutinee not in WHNF is used
again in an alternative, despite it no longer being available according to our system.

An expression in which a variable $x$ occurs both in the case scrutinee and in
the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
considered in isolation:
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
If $x$ refers to an unevaluated expression, then scrutinizing will force the expression
bound by $x$ to be evaluated to WHNF. In a subsequent use of
$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
must be consumed. Since $x$ is just another name (like the case binder) for the
scrutinee in WHNF, we may use it instead of the case binder or pattern
variables.
%
On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
scrutinizing it in a case expression is a no-op, thus we may use it again (in
mutual exclusion with the case binder and pattern variables).

However, allowing this expression to be linear makes $\lambda^\pi_\Delta$ unsound!
Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
where $y$ is a free linear variable, and assume $\lambda x.~\ccase{x}{\_ \to x}$
to be a linear function. According to $\lambda^\pi_\Delta$'s call-by-name
$\beta$-reduction, we have:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBName}
\ccase{use~y}{\_ \to use~y}
\end{array}
\]
After the reduction, $y$ is a linear variable \emph{consumed} in the scrutinee
of the case that also occurs in the case alternative -- the resource
$y$ is being effectively duplicated.
%
If the application were instead reduced with call-by-need
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBNeed}
\llet{x = use~y}{\ccase{x}{\_ \to x}}
\end{array}
\]
then the computation using $y$ would be let-bound, and $y$ would be used as a
scrutinee variable, which preserves semantic linearity.

For an optimising compiler such as GHC, the reverse binder swap transformation
violating linearity in Linear Core is relevant because optimising
compilers do not necessarily allocate a binding/use call-by-need for
every function application -- our result shows that failing to do so
when calling a function such as the one above is \emph{unsound}.
%
In an optimising compiler, a binding may be deemed unnecessary if the function
argument is inferred to be used exactly once. Linearity can play a
(non-heuristic) role in this inference, for a linear function uses its
argument exactly once by definition.
%
% Since the compiler may leverage this information (we
% briefly discuss \emph{linearity-influenced optimisations} in
% Section~\ref{sec:discussion}).
%
However, it is incompatible to both type expressions like $(\lambda
x.~\ccase{x}{\_ \to x})$ as linear functions and consider
call-by-name $\beta$-reduction of linear functions unconditionally
safe (without risk of duplicating resources nor work).
%
Furthermore, considering such an use of $x$ to be linear also makes inlining of
linear let bindings no longer unconditionally safe.

>>>>

When the scrutinee is not in WHNF, it is not statically known which resources
will have been consumed after evaluating said scrutinee, or which
are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
a \rightarrow e\}}$:
\begin{enumerate}

@@

    \item On the other hand, since evaluation to WHNF does not guarantee all
    resources are consumed (unless the result matches an
    unrestricted pattern), we must ensure that any resources that transfer from the
    scrutinee to the alternative are ultimately consumed (in the example, $x$
    is not consumed by evaluating the scrutinee for $f = K$).

@@

We develop the metatheory of Linear Core by first 
presenting the operational semantics of $\lambda^\pi_\D$,
consisting of a lazy natural semantics in the
style of Launchbury~\cite{10.1145/158511.158618}.
This semantics conveys the expected behavior of an implementation of
lazy evaluation and is agnostic to any linearity information.
Such a semantics is ill-suited for reasoning about
linearity and so we develop an \emph{instrumented} linearity-aware semantics that is
enriched with sufficient information to allow us to establish type
safety, thus showing that linearity is preserved. We show that the natural
and instrumented semantics are bisimilar for well-typed terms and so
derive type safety in the natural semantics
(Section~\ref{sec:typesafe}). Finally, we show that multiple optimising
transformations such as those performed by GHC on Linear Core are
type preserving
(Section~\ref{sec:optimisations-preserve-types-meta}), and thus
linearity preserving.

@@

The Launchbury-style natural semantics, presented in Figure~\ref{fig:opsem}, captures
standard lazy evaluation while ignoring linearity information.
% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide. 
%
%Both semantics are
The natural semantics is equipped with an evaluation environment that
maps variables to expressions and is
mutated in order to express evaluation\todo{aqui não seria sharing?}. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders (see Appendix~\ref{app:opsem}).
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.

@@

  The natural semantics is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, and $\Theta$ and $\Theta'$ are evaluation environments with bindings of
the form $x:_\omega\sigma = e$ which assigns the expression $e$ to the variable
$x$ of the given type $\s$. The rules are standard, augmenting the runtime
environment with let-bound expressions which are evaluated (only once) if the
corresponding let-bound variable is forced.

@@

While the natural semantics is a mostly direct model of an implementation
of lazy evaluation, it is inconvenient for formal reasoning about
well-typed linear terms.
%
We follow an approach
similar to that of \cite{cite:linearhaskell}, defining
an \emph{instrumented} operational semantics in which linear variables are
erased from 
the runtime environment once they are forced, ensuring that any term violating linearity
will result in a stuck state. The semantics is also type-aware and
carries with it sufficient data to reconstruct typing derivations for
the purposes of showing type safety.

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$\todo{isto não me fez sentido} which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.\todo{Na semantica O omega e 1 ainda estão separados, é assim?}

@@

%Finally, we discuss the special case of scrutinizing a
%single variable $x$, e.g.:
%$
%\lambda x.~\ccase{x}{\_ \to x}
%$. 
%It might seem as though this function is linear:
%%begin{itemize}
%% \item
%if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
%just returns the resource intact;
%% \item
%if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
%$x$ returns the result of evaluating $x$ that still had to be consumed exactly
%once.
%%\end{itemize}
%However, in practice, this program's linearity can only be determined by
%further analysis on the non-strict evaluation strategy. If linear
%function applications are $\beta$-reduced \emph{call-by-name}
%(often desirable, as linear functions are known to use their argument exactly
%once)
%% and thus call-by-name reduction will definitely not duplicate computation
%while the example function is considered linear, then such an application might
%duplicate linear resources during evaluation. For instance:
%\[
%\begin{array}{l}
%(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
%\Longrightarrow_{CBN}
%\ccase{free~x}{\_ \to free~x}
%\end{array}
%\]
%On the other hand, if linear function applications are reduced
%\emph{call-by-need}, the above example would instead be semantically linear
%because the result of freeing $x$ would be bound to some $y$ first.
%%
%The system we present in the next section does accept this function
%scrutinizing a variable as linear, as our operational semantics are
%call-by-need.
%This interaction between evaluation and linearity is considered specifically in the
%reverse binder swap transformation discussed in
%Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}
%

Reverse binder swap is dual to the binder swap transformation, substituting
occurrences of the case binder $z$ by the scrutinee in the special case when
the scrutinee is a variable $x$.
%
% his is dual to what the binder swap transformation
% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
By using the scrutinee $x$ instead of the case binder, we might be able to
float out expressions from the alternative using the case binder:
\[
    \ccase{x}{z~\{\rho \rightarrow e\}} \Rightarrow \ccase{x}{z~\{\rho \rightarrow e[x/z]\}}
\]

%For example, we might float an expensive computation involving $z$ out of the
%case alternative, where $z$ is now out of scope:
%
%{\small\[
%\begin{array}{l}
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
%\Longrightarrow_\textrm{Reverse binder swap}\\
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
%\Longrightarrow_\textrm{Commuting case-let}\\
%\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
%\end{array}
%\]}
%
%\noindent If $go$ is compiled into a loop, we only compute
%$\mathit{expensive}~x$ once instead of in every loop iteration.
%
Even though GHC applies the reverse binder swap transformation in the
Core-to-Core passes, this optimisation violates linearity in Linear Core
because the transformed program is rejected -- the scrutinee $x$, a single
variable, is not considered to be in WHNF by our system, thus occurrences of
$x$ in any alternative are ill-typed in it. This is a conservative choice in $\lambda^\pi_\Delta$ for
typing the subtle case where the scrutinee is a single variable:

Consider $\ccase{x}{\_ \to x}$. Perhaps surprisingly, this expression can be
considered semantically linear in a call-by-need calculus but violates
linearity in a call-by-name one:
%
%An expression in which a variable $x$ occurs both in the case scrutinee and in
%the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
%considered in isolation:
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
Under call-by-need, if $x$ refers to an unevaluated expression, scrutinizing it forces the expression
bound by $x$ to WHNF. In a subsequent use of
$x$ in an alternative, $x$ refers to the already evaluated scrutinee in WHNF.
Since $x$ is just another name (like the case binder) for the scrutinee in
WHNF, its use is also valid in the alternatives.
%
Dually, if $x$ refers to an expression already in WHNF, no evaluation takes
place and $x$ is still just another name for the scrutinee in WHNF.

However, under call-by-name, evaluation of $x$ is not shared:
%However, allowing this expression to be linear makes $\lambda^\pi_\Delta$ unsound!
%Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
%where $y$ is a free linear variable, and assume $\lambda x.~\ccase{x}{\_ \to x}$
%to be a linear function. According to $\lambda^\pi_\Delta$'s call-by-name
%$\beta$-reduction, we have:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBName}
\ccase{use~y}{\_ \to use~y}
\end{array}
\]
After the reduction, $y$ is a linear variable consumed both in the scrutinee
and in the case alternative -- $y$ has been duplicated!
%
If the application is instead reduced call-by-need, $y$ is not duplicated:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBNeed}
\llet{x = use~y}{\ccase{x}{\_ \to x}}
\end{array}
\]
%then the computation using $y$ would be let-bound, and $y$ would be used as a
%scrutinee variable, which preserves semantic linearity.

In an optimising compiler such as GHC, there is no definitive evaluation
strategy. It is mostly up to heuristics to determine what to inline
\emph{à la} call-by-name and what to create let-bindings for \emph{à
la} call-by-need.
%
In this flexible evaluation setting, we highlight the reverse binder swap is
only a linearity-preserving transformation as long as a linear variable with
more than one occurrence is never substituted for an expression with free
linear variables.
%
%This is unfortunately at odds with
%Moreover, allowing the reverse binder swap also challenges the 
%
%It additionally challenges the non-heuristic property that linear variables can
%be unconditionally inlined.......

%the aggressive inlining passes blur evaluation
%there is no definitive evaluation
%strategy in the sense functional compilers use both when agressively inline
%expressions \emph{à la} call-by-name.
%call-by-need for
%every function application -- our result shows that failing to do so
%when calling a function such as the one above is \emph{unsound}.
%
%In an optimising compiler, a binding may be deemed unnecessary if the function
%argument is inferred to be used exactly once. Linearity can play a
%(non-heuristic) role in this inference, for a linear function uses its
%argument exactly once by definition.
%
% Since the compiler may leverage this information (we
% briefly discuss \emph{linearity-influenced optimisations} in
% Section~\ref{sec:discussion}).
%
%However, it is incompatible to both type expressions like $(\lambda
%x.~\ccase{x}{\_ \to x})$ as linear functions and consider
%call-by-name $\beta$-reduction of linear functions unconditionally
%safe (without risk of duplicating resources nor work).
%
%Furthermore, considering such an use of $x$ to be linear also makes inlining of
%linear let bindings no longer unconditionally safe.
